---
title: "Comparison with read.dbc and microdatasus"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparison with read.dbc and microdatasus}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Several R packages provide access to DATASUS data. This article compares
`datasusr` with the two most established alternatives — `read.dbc` and
`microdatasus` — to help you choose the right tool for your workflow.

## Overview

| Feature | datasusr | read.dbc | microdatasus |
|---|---|---|---|
| Read DBC files | In-memory C parser | Temp DBF + `foreign::read.dbf` | Via `read.dbc` dependency |
| Column selection at read time | Yes (`select`) | No | Yes (`vars`) |
| Type control (`col_types`, `guess_types`) | Yes | No | No |
| Date parsing at read time | Yes (`parse_dates`) | No | Via `process_*()` post-hoc |
| Column names in snake\_case | Yes (`clean_names`) | No (uppercase) | No (uppercase) |
| Returns tibble | Always | `data.frame` | `tibble` (after processing) |
| FTP catalog / discovery | Yes (18 sources) | No | Yes (`fetch_datasus()`) |
| Download with caching | Yes (persistent, configurable) | No | Yes (session-level) |
| Parallel downloads with progress | Yes (`curl::multi_download`) | No | No |
| Variable labelling / recoding | No | No | Yes (`process_*()` functions) |
| One-step fetch + read | Yes (`datasus_fetch()`) | No | Yes (`fetch_datasus()`) |
| CRAN status | GitHub | Archived | Archived (dep on `read.dbc`) |
| License | MIT | AGPL-3 | MIT |

## DBC reading: architecture

The three packages take fundamentally different approaches to reading
`.dbc` files:

**read.dbc** decompresses the `.dbc` to a temporary `.dbf` on disk using C
code (`dbc2dbf`), then calls `foreign::read.dbf()` to parse it. This
two-step process is reliable but involves disk I/O for the intermediate
file and inherits the type limitations of `foreign::read.dbf()` (all
numeric fields become `double`; no column selection).

**microdatasus** depends on `read.dbc` for the file reading step. Its main
value is in the `fetch_datasus()` download function and the `process_*()`
family, which recode categorical variables (e.g. replacing municipality
codes with names, labelling sex, age groups, etc.). It also supports
column selection via the `vars` parameter.

**datasusr** performs decompression and DBF parsing entirely in memory in a
single C call. The compressed payload is decompressed with `blast` (the same
algorithm used by `read.dbc`) directly into a memory buffer, then parsed
field-by-field into pre-allocated R vectors. This avoids temporary files
and enables features like column selection (`select`), type inference
(`guess_types`), explicit type control (`col_types`), and date parsing
(`parse_dates`) at the C level.

## Download and catalog features

Both `datasusr` and `microdatasus` provide functions to discover and download
files from the DATASUS FTP. They differ in design:

**microdatasus** exposes a single `fetch_datasus()` function that takes
high-level parameters (`year_start`, `year_end`, `uf`,
`information_system`) and handles path construction internally. It downloads,
reads, and binds files in one call. This is convenient but offers limited
control over which specific files are retrieved.

**datasusr** offers a layered API: `datasus_sources()` and
`datasus_file_types()` let you browse the catalog, `datasus_list_files()`
generates and validates file lists, `datasus_download()` handles downloading
with caching, and `datasus_fetch()` wraps them all for one-step convenience.
This design lets you inspect and customise each step.

Both packages support caching of downloaded files. `datasusr` provides
a persistent, configurable cache with explicit management functions
(`datasus_cache_info()`, `datasus_cache_prune()`, `datasus_cache_clear()`)
and supports configuration via environment variable, R option, or function
argument.

## Variable labelling

One area where `microdatasus` stands out is **post-processing**. Its
`process_sim()`, `process_sinasc()`, `process_sih()`, `process_sia()`, and
`process_cnes()` functions decode categorical fields: they replace numeric
codes with human-readable labels for variables like sex, race, municipality,
cause of death (ICD-10), and procedures. This saves considerable manual
work when building descriptive analyses.

`datasusr` does not currently provide variable labelling. It focuses on
fast, flexible I/O and leaves the semantic interpretation to the analyst.
If you need labelled variables, you can apply your own lookup tables on the
tibble returned by `datasusr`.

## CRAN availability

As of this writing, both `read.dbc` and `microdatasus` have been archived
from CRAN. `microdatasus` was archived due to its dependency on `read.dbc`.
Both remain installable from GitHub. `datasusr` is available on GitHub and
has no dependency on either package.

## When to use which

Use **`datasusr`** when you need fast, flexible reads of `.dbc` files — for
example, reading specific columns from large files, controlling types, or
building reproducible pipelines with persistent caching. It works well as
a standalone package for data engineering tasks and integrates naturally
with the tidyverse.

Use **`microdatasus`** when you want labelled, analysis-ready data frames
with minimal code. The `process_*()` functions are a significant time-saver
for exploratory epidemiological analyses.

Use **`read.dbc`** when you just need to read a `.dbc` file into a data
frame and don't need column selection, type control, or download helpers.

## Code comparison

Reading a single DBC file with each package:

```{r eval = FALSE}
# --- datasusr -----------------------------------------------------------------
library(datasusr)
x <- read_datasus_dbc(
  "RDPE2401.dbc",
  select = c("uf_zi", "ano_cmpt", "val_tot"),
  col_types = c(val_tot = "double")
)

# --- read.dbc -----------------------------------------------------------------
library(read.dbc)
x <- read.dbc("RDPE2401.dbc")        # returns full data.frame, all columns uppercase
x <- x[, c("UF_ZI", "ANO_CMPT", "VAL_TOT")]

# --- microdatasus -------------------------------------------------------------
library(microdatasus)
x <- fetch_datasus(
  year_start = 2024, year_end = 2024,
  month_start = 1, month_end = 1,
  uf = "PE", information_system = "SIH-RD"
)
x <- process_sih(x)
```

Downloading and reading a full year:

```{r eval = FALSE}
# --- datasusr -----------------------------------------------------------------
df <- datasus_fetch(
  source = "SIHSUS", file_type = "RD",
  year = 2024, month = 1:12, uf = "PE",
  select = c("uf_zi", "ano_cmpt", "val_tot")
)

# --- microdatasus -------------------------------------------------------------
df <- fetch_datasus(
  year_start = 2024, year_end = 2024,
  uf = "PE", information_system = "SIH-RD"
)
df <- process_sih(df)
```
